Assessing the Firing of Ceramic Materials: a Seriation-Based Approach

### 2023-12-04 analysis/paper/paper.Rmd
Use Rmarkdown instead of Quarto

---
title: "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
author:
  - Nicolas Frerebeau:
      correspondence: "yes"
      email: nicolas.frerebeau@u-bordeaux-montaigne.fr
      orcid: 0000-0001-5759-4944
      affiliations:
        - name: "UMR 6034 Archéosciences Bordeaux"
          address: "Maison de l'Archéologie, Université Bordeaux Montaigne"
          city: Pessac cedex
          country: France
          postal-code: 33607
          url: https://www.archeosciences-bordeaux.fr
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    include-in-header: 
      - text: |
          \usepackage{booktabs}
    classoption: 
      - french
      - 12pt
      - a4paper
      - oneside
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    comment: "#>"
    fig.path: "../figures/"
    fig.dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: [references.bib, packages.bib]
csl: "../templates/archeosciences.csl"
abstract: |
  The study of the firing of archaeological ceramics is a key factor in 
  interpreting ancient techniques and related social practices. This study 
  proposes a new method based on the correspondence analysis of powder X-ray 
  diffraction data (XRD), aiming to provide an alternative to the so-called
  Equivalent Firing Temperature framework. The properties of correspondence 
  analysis and its application for matrix seriation enable the description of 
  the firing of a ceramic assemblage. The application of correspondence 
  analysis to XRD data of Iron Age ceramics revealed the  significant potential 
  of this statistical method in the technological analysis of ancient pottery. 
  This allows to order a large number of samples according to a firing 
  gradient and to track mineralogical transformations occurring during firing.
  This method should let to more precise comparisons than those based on 
  estimating the equivalent firing temperature.
keywords: |
  X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
---

```{r}
#| label: setup
options(knitr.kable.NA = "")
```

@@

| **Version** | **Date**   | **Description**            | **DOI** |
|---------:|-----------:|:------------------------------|:---------------------------|
| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020138](https://doi.org/10.5281/zenodo.8020138) |
| 1.1      | 2023-08-01 | Corrections based on reviews  | [10.5281/zenodo.8204409](https://doi.org/10.5281/zenodo.8204409) |

@@

`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop (Foz-Calanda, Teruel, Aragon) were sampled to represent the entire period of the workshop's activity (@tbl-corpus). The workshop was active between 225-200 BC and 40-30 AD. Previous archaeological studies of the workshop have highlighted three successive phases of activity, characterized by changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012]. Several kilns have been excavated, with discontinuation and destruction occurring around 50-40 BC [@gorgues2012]. The use of a larger amphora kiln, the production of Roman amphorae, and the presence of Latin epigraphy indicate a major reorganization of the workshop in the middle of the 1st century. The workshop ceased its activity in the following decade due to territorial reorganization under Roman influence and its distance from trade routes [@gorgues2012]. This workshop provides a context well-suited for investigating the technological choices related to ceramic firing in the Ebro Valley at the end of the Iron Age.

@@

A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017]. Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here (@tbl-corpus).

Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled (@tbl-corpus). These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dated between the 8^th^ and the 4^th^ century BC). Comparisons with these two settlements are not intended to establish genetic relationships (in a provenance meaning), but rather to propose a comparative technology study. Samples from El Palao and Torre Cremada have the same compositional ranges in the CaO-Al~2~O~3~-SiO~2~ system and belong to the same category of Iberian fine wares as the material from Mas de Moreno (see below).

```{r}
#| label: tbl-corpus
#| tbl-cap: "Ceramic and clay (unfired pottery sherds) samples analyzed in this study. Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC. LOI: loss on ignition [@frerebeau2020a; @frerebeau2015g; @frerebeau2015e]."
corpus |> 
  merge(y = xrf_cas[, -2], by = "sample") |> 
  (function(x) x[order(x$sample), ])() |> 
  (function(x) x[, !(colnames(x) %in% c("site_type", "stratigraphy", "artefact", "decoration", "part"))])() |> 
  (function(x) {
    write.table(x, file = here::here("analysis/tables/corpus.csv"), 
                col.names = TRUE, row.names = FALSE, sep = ",", dec = ".")
    x
  })() |> 
  knitr::kable(
    col.names = c("Sample", "Material", "Site", "Phase", "LOI (%)", "CaO (%)", "SiO2 (%)", "Al2O3 (%)"),
    booktabs = TRUE,
    longtable = TRUE
  )
```

@@

The results of the correspondence analysis of the diffractograms are presented in @fig-plot-ca. Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and sherds. The data from El Palao and Torre Cremada were introduced into the analysis as supplementary individuals. These additional individuals do not contribute to the calculation of the correspondence analysis itself, but they are projected onto the existing analysis space to visualize their relationships with the original dataset.

The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (@fig-plot-ca A and B). Plotting the coordinates of the individuals shows an interesting pattern (@fig-plot-ca C). The upper quadrants each consist of a small group of closely located individuals, likely exhibiting high similarity to one another. The remaining samples are distributed in the lower quadrants.

@@

@fig-plot-contrib shows the contribution of the different columns to the construction of the first two CA-axes. It is important to note that the analysis is performed on the diffractograms: a diffraction peak represents a range of $2\theta$ positions, spanning multiple columns in the data matrix. Consequently, the columns that contribute the most to the construction of the first axis (with a contribution greater than 0.5) are associated with the following diffraction peaks: 21.94° (plagioclases), 26.64° (quartz), 27.76° (plagioclases), 27.90° (plagioclases), and 29.42° (calcite). Similarly, for the second axis, the columns that contribute the most are those corresponding to the peaks at 12.32° (phyllosilicate), 26.66° (quartz), and 31.36° (melilites).

@@

The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (@fig-plot-ca, @fig-plot-guttman). This effect is characterized by a strong nonlinear relationship between the second factor and the first factor. The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014]. Interestingly, the samples located in the upper left quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A, @fig-plot-ca D, @fig-plot-xrd). These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1. Conversely, the samples located in the two lower quadrants are linked to the peaks of quartz and the melilite series.

@@

```{r}
#| label: fig-plot-xrd
#| fig-cap: "Diffractograms of representative samples from each quadrant of the CA plot (see @fig-plot-ca). From top to bottom: upper left quadrant (BDX14922), lower left quadrant (BDX15451), lower right quadrant (BDX15465), upper right quadrant (BDX22203). Only the peaks that contribute the most to the construction of the first two CA axes are indexed (see text and @fig-plot-contrib)."
#| fig-width: 10
#| fig-height: 11
par(mfrow = c(4, 1), mar = c(1, 0, 1, 1), oma = c(4, 4, 4, 0))

@@

The principal axis of correspondence analysis is built upon the opposition between primary phases and secondary phases. The resulting projection highlights the progression of firing, in accordance with the expected mineralogical transformations (@fig-plot-ca D). The intermediate position of gehlenite can also be explained. The ceramics from Mas de Moreno all belong to the category of Ca-rich ceramics, as defined by @noll1991 [@frerebeau2020a]: within the CaO-Al~2~O~3~-SiO~2~ system, they fall within the quartz-anorthite-wollastonite triangle (@fig-plot-guttman D). In this case, the formed gehlenite is metastable. It serves as an intermediate product towards the formation of anorthite and wollastonite, linked to the diffusion of calcium through the matrix [@traore2000; @traore2003a]. As gehlenite forms a solid solution with akermanite (the melilite series), it can also promote the crystallization of diopside (primarily of the fassaite type) rather than anorthite [@dondi1998; @rathossi2010].

@@

Interestingly, @fig-plot-loi-map (top) illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples [see @frerebeau2020a for the LOI data]. It is observed that the LOI is inversely correlated with the coordinates of the individuals along the first axis. To further explore this relationship, a second-order polynomial regression analysis was conducted. The results indicated that the overall regression was statistically significant (RSE = `r round(loi_fit_res$sigma, 2)` on `r  loi_fit_res$df[2]` degrees of freedom, R^2^ = `r round(loi_fit_res$r.squared, 2)`, adjusted R^2^ = `r round(loi_fit_res$adj.r.squared, 2)`, `r sprintf("F(%g, %g) = %g", loi_fit_res$fstatistic[2], loi_fit_res$fstatistic[3], round(loi_fit_res$fstatistic[1], 1))`, p < 0.001), indicating a good fit of the polynomial regression model to the data.

Therefore, it appears that the ceramics from Mas de Moreno can be ordered according to the progress of these mineralogical transformations or, equivalently, according to the energy received during the firing (@fig-plot-loi-map, bottom).

@@

The material from the Mas de Moreno is the result of waste from the production process. It is therefore necessary to distinguish between what is representative of material produced in the workshop and what is the result of a non-desirable situation leading to deliberate rejection (firing defect). In order to explore these two hypotheses, the diffractograms of several artifacts from consumption contexts (El Palao and Torre Cremada) were introduced into the previous analysis as supplementary individuals. This was made possible by the technical homogeneity of the material studied. Samples from El Palao and Torre Cremada have the same compositional ranges (@fig-plot-guttman D) and belong to the same category of Iberian fine wares as the material from Mas de Moreno.

These additional individuals are all situated in the lower section of the first plane of the correspondence analysis (@fig-plot-guttman C). This enables the differentiation of the artifacts from Mas de Moreno into those associated with regular production and those linked to misfiring (i.e., individuals located in the upper right quadrant; @fig-plot-guttman A, @fig-plot-ca D, @fig-plot-loi-map bottom). Individuals reflecting a typical production pattern exhibit peaks related to gehlenite rather than anorthite and diopside. This suggests a certain level of control over the energy input during firing to allow the development of this intermediate phase.

@@

Finally, it is worth noting that there appears to be no relationship with chronology (@fig-plot-guttman B). Despite significant workshop modifications, including the adoption of a large amphora kiln that resulted in a shift in production quantities during the final years of the workshop's operation [@gorgues2012], the produced artefacts exhibit consistent mineralogical characteristics.

>>>>

---
title: "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
author:
  - Nicolas Frerebeau:
      correspondence: "yes"
      email: nicolas.frerebeau@u-bordeaux-montaigne.fr
      orcid: 0000-0001-5759-4944
      institute:
        - AB
institute:
  - AB:
      name: UMR 6034 Archéosciences Bordeaux
      address: Maison de l'Archéologie, Université Bordeaux Montaigne, 33607 Pessac cedex, France
date: "Accepted for publication on 2023-09-15"
bibliography: [references.bib, packages.bib]
csl: "../templates/archeosciences.csl"
abstract: |
  The study of the firing of archaeological ceramics is a key factor in 
  interpreting ancient techniques and related social practices. This study 
  proposes a new method based on the correspondence analysis of powder X-ray 
  diffraction data (XRD), aiming to provide an alternative to the so-called
  Equivalent Firing Temperature framework. The properties of correspondence 
  analysis and its application for matrix seriation enable the description of 
  the firing of a ceramic assemblage. The application of correspondence 
  analysis to XRD data of Iron Age ceramics revealed the  significant potential 
  of this statistical method in the technological analysis of ancient pottery. 
  This allows to order a large number of samples according to a firing 
  gradient and to track mineralogical transformations occurring during firing.
  This method should let to more precise comparisons than those based on 
  estimating the equivalent firing temperature.
keywords: |
  X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
output: 
  bookdown::pdf_document2:
    toc: false
    fig_caption: yes
    number_sections: true
    pandoc_args:
    - --lua-filter=../templates/scholarly-metadata.lua
    - --lua-filter=../templates/author-info-blocks.lua
    - --lua-filter=../templates/pagebreak.lua
---

```{r}
#| label: setup
#| include: false
options(knitr.kable.NA = "")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  collapse = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  fig.dpi = 600
)
```

@@

| **Version** | **Date**   | **Description**            | **DOI** |
|---------:|-----------:|:------------------------------|:---------------------------|
| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020138](https://doi.org/10.5281/zenodo.8020138) |
| 1.1      | 2023-08-01 | Corrections based on reviews  | [10.5281/zenodo.8204409](https://doi.org/10.5281/zenodo.8204409) |
| 1.2      | 2023-09-15 | Accepted for publication      |  |

@@

`r sum(corpus$material == "ceramic" & corpus$site == "Mas de Moreno")` ceramics from the Mas de Moreno workshop (Foz-Calanda, Teruel, Aragon; table \@ref(tab:tbl-corpus)) were sampled to represent the entire period of the workshop's activity [@frerebeau2020a; @frerebeau2015g]. The workshop was active between 225-200 BC and 40-30 AD. Previous archaeological studies of the workshop have highlighted three successive phases of activity, characterized by changes in the productive space and the emergence of Roman elements [@gorgues2007; @gorgues2012]. Several kilns have been excavated, with discontinuation and destruction occurring around 50-40 BC [@gorgues2012]. The use of a larger amphora kiln, the production of Roman amphorae, and the presence of Latin epigraphy indicate a major reorganization of the workshop in the middle of the 1st century. The workshop ceased its activity in the following decade due to territorial reorganization under Roman influence and its distance from trade routes [@gorgues2012]. This workshop provides a context well-suited for investigating the technological choices related to ceramic firing in the Ebro Valley at the end of the Iron Age.

@@

A number of fragments of unfired potteries were also found during the excavation of the workshop [for a complete presentation of these unfired potsherds, see @frerebeau2017; @frerebeau2015e]. Of these, `r sum(corpus$material == "clay" & corpus$site == "Mas de Moreno")` samples have been analysed here (table \@ref(tab:tbl-corpus)).

Finally, `r sum(corpus$material == "ceramic" & corpus$site != "Mas de Moreno")` additional Iberian fine wares were sampled (table \@ref(tab:tbl-corpus)). These came from two settlements close to the workshop: Torre Cremada (Valdeltormo, Teruel, Aragon; `r sum(corpus$site == "Torre Cremada")` samples from the 1^st^ century BC) and El Palao (Alcañiz, Teruel, Aragon; `r sum(corpus$site == "El Palao")` samples dated between the 8^th^ and the 4^th^ century BC). Comparisons with these two settlements are not intended to establish genetic relationships (in a provenance meaning), but rather to propose a comparative technology study. Samples from El Palao and Torre Cremada have the same compositional ranges in the CaO-Al~2~O~3~-SiO~2~ system and belong to the same category of Iberian fine wares as the material from Mas de Moreno (see below).

```{r}
#| label: tbl-corpus
corpus |> 
  merge(y = xrf_cas[, -2], by = "sample") |> 
  (function(x) x[order(x$sample), ])() |> 
  (function(x) x[, !(colnames(x) %in% c("site_type", "stratigraphy", "artefact", "decoration", "part"))])() |> 
  (function(x) {
    write.table(x, file = here::here("analysis/tables/corpus.csv"), 
                col.names = TRUE, row.names = FALSE, sep = ",", dec = ".")
    x
  })() |> 
  knitr::kable(
    col.names = c("Sample", "Material", "Site", "Phase", "LOI (%)", "CaO (%)", "SiO2 (%)", "Al2O3 (%)"),
    caption = "Ceramic and clay (unfired pottery sherds) samples analyzed in this study. Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC. LOI: loss on ignition.",
    booktabs = TRUE,
    longtable = TRUE
  )
```

@@

The results of the correspondence analysis of the diffractograms are presented in fig. \@ref(fig:fig-plot-ca). Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and sherds. The data from El Palao and Torre Cremada were introduced into the analysis as supplementary individuals. These additional individuals do not contribute to the calculation of the correspondence analysis itself, but they are projected onto the existing analysis space to visualize their relationships with the original dataset.

The first two dimensions are sufficient to retain about `r round(xrd_moreno_eig[2, 3])`% of the total inertia contained in the data (fig. \@ref(fig:fig-plot-ca)A and B). Plotting the coordinates of the individuals shows an interesting pattern (fig. \@ref(fig:fig-plot-ca)C). The upper quadrants each consist of a small group of closely located individuals, likely exhibiting high similarity to one another. The remaining samples are distributed in the lower quadrants.

@@

fig. \@ref(fig:fig-plot-contrib) shows the contribution of the different columns to the construction of the first two CA-axes. It is important to note that the analysis is performed on the diffractograms: a diffraction peak represents a range of $2\theta$ positions, spanning multiple columns in the data matrix. Consequently, the columns that contribute the most to the construction of the first axis (with a contribution greater than 0.5) are associated with the following diffraction peaks: 21.94° (plagioclases), 26.64° (quartz), 27.76° (plagioclases), 27.90° (plagioclases), and 29.42° (calcite). Similarly, for the second axis, the columns that contribute the most are those corresponding to the peaks at 12.32° (phyllosilicate), 26.66° (quartz), and 31.36° (melilites).

@@

The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (fig. \@ref(fig:fig-plot-ca), \@ref(fig:fig-plot-guttman)). This effect is characterized by a strong nonlinear relationship between the second factor and the first factor. The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014]. Interestingly, the samples located in the upper left quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (fig. \@ref(fig:fig-plot-guttman)A, \@ref(fig:fig-plot-ca)D, \@ref(fig:fig-plot-xrd)). These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1. Conversely, the samples located in the two lower quadrants are linked to the peaks of quartz and the melilite series.

@@

```{r}
#| label: fig-plot-xrd
#| fig-cap: "Diffractograms of representative samples from each quadrant of the CA plot (see fig. \\@ref(fig:fig-plot-ca)). From top to bottom: upper left quadrant (BDX14922), lower left quadrant (BDX15451), lower right quadrant (BDX15465), upper right quadrant (BDX22203). Only the peaks that contribute the most to the construction of the first two CA axes are indexed (see text and fig. \\@ref(fig:fig-plot-contrib))."
#| fig-width: 10
#| fig-height: 11
par(mfrow = c(4, 1), mar = c(1, 0, 1, 1), oma = c(4, 4, 4, 0))

@@

The principal axis of correspondence analysis is built upon the opposition between primary phases and secondary phases. The resulting projection highlights the progression of firing, in accordance with the expected mineralogical transformations (fig. \@ref(fig:fig-plot-ca)D). The intermediate position of gehlenite can also be explained. The ceramics from Mas de Moreno all belong to the category of Ca-rich ceramics, as defined by @noll1991 [@frerebeau2020a]: within the CaO-Al~2~O~3~-SiO~2~ system, they fall within the quartz-anorthite-wollastonite triangle (fig. \@ref(fig:fig-plot-guttman)D). In this case, the formed gehlenite is metastable. It serves as an intermediate product towards the formation of anorthite and wollastonite, linked to the diffusion of calcium through the matrix [@traore2000; @traore2003a]. As gehlenite forms a solid solution with akermanite (the melilite series), it can also promote the crystallization of diopside (primarily of the fassaite type) rather than anorthite [@dondi1998; @rathossi2010].

@@

Interestingly, figure \@ref(fig:fig-plot-loi-map) (top) illustrates the relationship between the first axis of the correspondence analysis and the loss on ignition (LOI) of the samples [see @frerebeau2020a for the LOI data]. It is observed that the LOI is inversely correlated with the coordinates of the individuals along the first axis. To further explore this relationship, a second-order polynomial regression analysis was conducted. The results indicated that the overall regression was statistically significant (RSE = `r round(loi_fit_res$sigma, 2)` on `r  loi_fit_res$df[2]` degrees of freedom, R^2^ = `r round(loi_fit_res$r.squared, 2)`, adjusted R^2^ = `r round(loi_fit_res$adj.r.squared, 2)`, `r sprintf("F(%g, %g) = %g", loi_fit_res$fstatistic[2], loi_fit_res$fstatistic[3], round(loi_fit_res$fstatistic[1], 1))`, p < 0.001), indicating a good fit of the polynomial regression model to the data.

Therefore, it appears that the ceramics from Mas de Moreno can be ordered according to the progress of these mineralogical transformations or, equivalently, according to the energy received during the firing (fig. \@ref(fig:fig-plot-loi-map), bottom).

@@

The material from the Mas de Moreno is the result of waste from the production process. It is therefore necessary to distinguish between what is representative of material produced in the workshop and what is the result of a non-desirable situation leading to deliberate rejection (firing defect). In order to explore these two hypotheses, the diffractograms of several artifacts from consumption contexts (El Palao and Torre Cremada) were introduced into the previous analysis as supplementary individuals. This was made possible by the technical homogeneity of the material studied. Samples from El Palao and Torre Cremada have the same compositional ranges (fig. \@ref(fig:fig-plot-guttman)D) and belong to the same category of Iberian fine wares as the material from Mas de Moreno.

These additional individuals are all situated in the lower section of the first plane of the correspondence analysis (fig. \@ref(fig:fig-plot-guttman)C). This enables the differentiation of the artifacts from Mas de Moreno into those associated with regular production and those linked to misfiring (i.e., individuals located in the upper right quadrant; fig. \@ref(fig:fig-plot-guttman)A, \@ref(fig:fig-plot-ca)D, \@ref(fig:fig-plot-loi-map) bottom). Individuals reflecting a typical production pattern exhibit peaks related to gehlenite rather than anorthite and diopside. This suggests a certain level of control over the energy input during firing to allow the development of this intermediate phase.

@@

Finally, it is worth noting that there appears to be no relationship with chronology (\@ref(fig:fig-plot-guttman)B). Despite significant workshop modifications, including the adoption of a large amphora kiln that resulted in a shift in production quantities during the final years of the workshop's operation [@gorgues2012], the produced artefacts exhibit consistent mineralogical characteristics.
